The zoon R package for reproducible and shareable species distribution modelling

### 2016-12-21 ms.Rmd
first crack at example 3

#### Example 3. Evaluating MaxLike

@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].

In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.

For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.

<!-- Possible experiments:
* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
* fitting the same models with GBIF data and evaluating against the BBS PA data,
* fitting/evaluating on spatially-stratified holdout data
* fitting the models with other species
  (choose only one of these)
  -->

All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.

What the workflow looks like:

```{r merowworkflowtemp, dpi = 600, eval = TRUE}
Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Background(n = 10000),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
```

```{r merowworkflow,dpi = 600, eval = FALSE}

Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Chain(Transform(trans = 'square',
                                 replace = FALSE), 
                       Background(n = 10000)),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
```

So we ran it again with disc-based spatial stratification:

```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
Merow_Silander_Spatial <- ChangeWorkflow(
    merow_and_silander,
    process = Chain(..., PartitionDisc))
```

>>>>

#### Example 3. What makes MaxEnt effective

MaxEnt is one of the most popular models for presence-only SDM [@Ahmed2015], in part because it performed very well in the comprehensive comparison of @Elith2006.
Unlike most other models use in SDM, MaxEnt considers occurrence records as arising from some probability distribution over space and attempts to model that distribution, subject to a 'maximum entropy' constraint [@Phillips2006].
Subsequent research has shown that although it approaches the problem from a diffent perspective, the model MaxEnt fits is in fact equivalent to Poisson point process model [@Renner?].
Moreover, so long as the number of background points large, this model is essentially the same as a logistic regression model [@?].
Given MaxEnt's treatment of the data is not fundamentally different than other common models, @Renner2013 and others have suggested that MaxEnt's predictive power is instead due to the array of features it constructs, and the fact that it applies regularisation when fitting the model.
Whilst the various features (candidate covariate transformations) increase the potential flexibility of the model, regularisation acts to prevent MaxEnt from overfitting to the data.

Here we illustrate how zoon can facilitate SDM methodological comparisons by fitting MaxEnt models with and without regularisation and feature construction.
We fit MaxEnt with both the usual java implementation, and an implementation recently provided in the `maxnet` R package [@maxnet] which approximates the full MaxEnt model via a regularised logistic regression.

The following zoon workflow fits five different models to a presence-background dataset, plots a prediciton map generated by each and evaluates the predictive performance of each model against a presence-absence dataset.
The presence-only and presence-absence data and covariates are for the Caroline wren in the USA, as analysed by @Royle2012 and @Merow2014, and distributed via the `maxlike` R package [@maxlike].
The five models compared are: the java implementation of MaxEnt, the `maxnet` implementation (via the `MaxNet` zoon module), the `maxnet` version without regularisation, without features, and without regularisation or features. 
As in @Merow2014, each model is trained on the same randomly selected subset of 500 occurrence records, with 10,000 randomly selected background records.
Statistics of the predictive performance of each fitted model are then calculated against the full presence-absence dataset, and prediction maps plotted in Figure 3.

```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE}
maxnet_comparison <- workflow(
  occurrence = CarolinaWrenPO,
  covariate  = CarolinaWrenRasters,
  process    = Chain(SubsampleOccurrence(500),
                     Background(n = 10000),
                     CarolinaWrenValidation),
  model      = list(MaxEnt(args = 'threshold=false'),
                    MaxNet,
                    MaxNet(regmult = 0.005),
                    MaxNet(features = 'l'),
                    MaxNet(regmult = 0.005, features = 'l')),
  output     = Chain(PerformanceMeasures,
                     PrintMap(points = FALSE)))
```

```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 9, fig.width = 1.5}
# plot the figure to disk, then plot in a separate chunk with a legend

par(mfrow = c(5, 1))

```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "150px", fig.cap = ""}
# knitr::include_graphics("figs/maxnet_plotting-1.png")
```

```{r maxent_aucs, eval = TRUE, echo = FALSE}
# AUCs, not quite sure what to do with these - embed them in the text somehow?
perflist <- maxnet_comparison$report[seq(1, 10, by = 2)]
AUCs <- sapply(perflist, function(x) x$auc)
names(AUCs) <- c('MaxEnt',
                 'MaxNet',
                 'MaxNet (no regularization)',
                 'MaxNet (no features)',
                 'MaxNet (no regularization or features)')
AUCs <- round(AUCs, digits = 4)
```

The AUC statistics calculated against the presence-absence data were:
```{r results = 'asis', echo = FALSE}
cat(paste(names(AUCs), AUCs, collapse = '; '))
cat('.')
```
Readers interested in other statistics, viewing response curves or adjusting the calidation method can easily reproduce and modify the analysis either by re-running the above code block or by downloading the workflow object from: <URL TBD>
